---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ggfips *is experimental* 🤩

... and most of the development work lives right here in the README. 



<!-- badges: start -->
<!-- badges: end -->

The goal of ggfips is to provide an interface where you can quickly build a map with a simple dataframe with fips codes and variables with characteristics.  geom_sf_fips brings the fips polygons (geometry vector) that you'll need to create maps based county and characteristics.

It uses Claus Wilke's fips geometries.  


# The status quo

```{r}
library(tidyverse)
library(sf)
#> Linking to GEOS 3.8.1, GDAL 3.2.1, PROJ 7.2.1

fips_geometries <- readRDS(url("https://wilkelab.org/SDS375/datasets/US_counties.rds")) %>%
  rename(FIPS = GEOID) %>% 
  janitor::clean_names()

US_census <- read_csv("https://wilkelab.org/SDS375/datasets/US_census.csv",
                      col_types = cols(FIPS = "c")
          ) %>% 
  janitor::clean_names()

# from Claus Wilke on ggplot2
fips_geometries %>%
  left_join(US_census, by = "fips") %>%
  ggplot() +
  geom_sf(aes(fill = mean_work_travel, 
              fips = fips), # a bug or a feature?
          linewidth = .1) + 
  scale_fill_viridis_c(option = "magma") ->
classic_plot_sf_layer

classic_plot_sf_layer
```



```{r}
layer_data(classic_plot_sf_layer) %>% 
  select(fips, geometry, xmin, xmax, ymin, ymax) ->
reference_fips

reference_fips %>% 
  left_join(fips_geometries) ->
reference_fips_full


save(reference_fips, file = "data/reference_fips.rda")
```



```{r}
compute_county <- function(data, scales){
 
  data %>% 
    # inner_join(fips_ggplot2_reference, multiple = "all") 
    inner_join(reference_fips) %>% 
    mutate(group = -1)

}


```


# Step 2: pass to ggproto

```{r}
StatCounty <- ggplot2::ggproto(`_class` = "StatCounty",
                                  `_inherit` = ggplot2::Stat,
                                  # required_aes = c("fips"), #breaks when required!?
                                  # setup_data = my_setup_data,
                                  compute_panel = compute_county,
                                  default_aes = aes(geometry = after_stat(geometry))
                                  )
```


# Step 3: write geom_* function

```{r}
geom_sf_county <- function(
  mapping = NULL,
  data = NULL,
  position = "identity",
  na.rm = FALSE,
  show.legend = NA,
  inherit.aes = TRUE, ...) {
  ggplot2::layer(
    stat = StatCounty,  # proto object from step 2
    geom = ggplot2::GeomSf,  # inherit other behavior
    data = data,
    mapping = mapping,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(na.rm = na.rm, ...)
  )
}
```

# step 4: test geom_* function, celebrate!

```{r, eval = T}
us_census <- read_csv("https://wilkelab.org/SDS375/datasets/US_census.csv",
                      col_types = cols(FIPS = "c")
          )  

head(us_census)

stamp_sf_county_usa <- function(fill = "darkgrey"){
  
  
  geom_sf_county(data = reference_fips %>% 
                   select(fips), 
                 fill = fill, 
                 aes(fips = fips))
  
}

us_census %>% 
  ggplot() + 
  aes(fips = FIPS) +
  stamp_sf_county_usa() +
  geom_sf_county(linewidth = .02, 
                      color = "darkgrey") +
  aes(fill = mean_work_travel) + 
  coord_sf() +
  scale_fill_viridis_c(option = "magma")
```


---


# Spotlighting

## can we make ggfips go further? Prototyping w/ Texas

```{r}
stamp_sf_county_texas <- function(fill = "darkgrey"){
  
  
  geom_sf_county(data = fips_geometries %>% filter(name_2 == "Texas") %>% 
                   select(fips = fips), 
                 fill = fill, 
                 aes(fips = fips))
  
}

  
reference_fips_full %>% filter(name_2 == "Texas") %>% 
    select(fips, geometry, xmin,xmax, ymin, ymax) ->
reference_fips_texas

compute_county_texas <- function(data, scales){
 
  data %>% 
    inner_join(reference_fips_texas) %>% 
    mutate(group = -1)

}


StatCountytexas <- ggplot2::ggproto(`_class` = "StatCountytexas",
                                  `_inherit` = ggplot2::Stat,
                                  # required_aes = c("fips"), #breaks when required!?
                                  # setup_data = my_setup_data,
                                  compute_panel = compute_county_texas,
                                  default_aes = aes(geometry = after_stat(geometry))
                                  )




geom_sf_county_texas <- function(
  mapping = NULL,
  data = NULL,
  position = "identity",
  na.rm = FALSE,
  show.legend = NA,
  inherit.aes = TRUE, ...) {
  ggplot2::layer(
    stat = StatCountytexas,  # proto object from step 2
    geom = ggplot2::GeomSf,  # inherit other behavior
    data = data,
    mapping = mapping,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(na.rm = na.rm, ...)
  )
}


stamp_sf_county_texas_gillespie <- function(fill = alpha("red", .3), 
                                            color = "red",
                                            alpha = .2, linewidth = 1){
  
  
  geom_sf_county(data = fips_geometries %>% filter(name_2 == "Texas") %>%
                   filter(name == "Gillespie") %>% 
                   select(fips = fips), 
                 fill = fill, color = color, alpha = alpha, linewidth = linewidth,
                 aes(fips = fips))
  
}



us_census %>% 
  ggplot() + 
  coord_sf() +
  aes(fips = FIPS) +
  stamp_sf_county_texas() +
  geom_sf_county_texas() +
  stamp_sf_county_texas_gillespie(alpha = 0, color = "green") +
  aes(fill = mean_work_travel) + 
  scale_fill_viridis_c(option = "magma")

```



---

# Generalize W colorado

I thought about providing convenience functions for every single county but that seemed like too much!  

```{r}
stamp_sf_county_state <- function(fill = "darkgrey", state = "Colorado"){
  
  state %>% tolower() -> state
  
  geom_sf_county(data = fips_geometries %>% 
                   mutate(name_2 = tolower(name_2)) %>% 
                   filter(name_2 == state) %>% 
                   select(fips = fips), 
                 fill = fill, 
                 aes(fips = fips))
  
}

  


compute_county_state <- function(data, scales, state = "Colorado"){
 
    state %>% tolower() -> state

  
  reference_fips_full %>% 
                   mutate(name_2 = tolower(name_2)) %>% 
                   filter(name_2 == state) %>% 
    select(fips, geometry, xmin,xmax, ymin, ymax) ->
reference_fips_state
  
  data %>% 
    inner_join(reference_fips_state) %>% 
    mutate(group = -1)

}


StatCountystate <- ggplot2::ggproto(`_class` = "StatCountystate",
                                  `_inherit` = ggplot2::Stat,
                                  # required_aes = c("fips"), #breaks when required!?
                                  # setup_data = my_setup_data,
                                  compute_panel = compute_county_state,
                                  default_aes = aes(geometry = after_stat(geometry))
                                  )




geom_sf_county_state <- function(
  mapping = NULL,
  data = NULL,
  position = "identity",
  na.rm = FALSE,
  show.legend = NA,
  inherit.aes = TRUE, ...) {
  ggplot2::layer(
    stat = StatCountystate,  # proto object from step 2
    geom = ggplot2::GeomSf,  # inherit other behavior
    data = data,
    mapping = mapping,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(na.rm = na.rm, ...)
  )
}


stamp_sf_county_state_county <- function(state = "Colorado",
                                         county = "Arapahoe",
                                         fill = alpha("red", .3),
                                            color = "red",
                                            alpha = .2, linewidth = 1){

      state %>% tolower() -> state
    county %>% tolower() -> county


  geom_sf_county(data = fips_geometries %>% 
                   mutate(name_2 = tolower(name_2)) %>% 
                   filter(name_2 == state) %>%
                   mutate(name = tolower(name)) %>% 
                   filter(name == county) %>%
                   select(fips = fips),
                 fill = fill, color = color, alpha = alpha, linewidth = linewidth,
                 aes(fips = fips))

}



us_census %>% 
  ggplot() + 
  coord_sf() +
  aes(fips = FIPS) +
  stamp_sf_county_state(state = "colorado") +
  geom_sf_county_state(state = "colorado") +
  stamp_sf_county_state_county(alpha = 0, color = "green") +
  aes(fill = mean_work_travel) + 
  scale_fill_viridis_c(option = "magma")

us_census %>% 
  ggplot() + 
  coord_sf() +
  aes(fips = FIPS) +
  geom_sf_county_state(state = "new york") +
  stamp_sf_county_state_county(alpha = 0, color = "green", state = "new york", county = "westchester") +
  aes(fill = mean_work_travel) + 
  scale_fill_viridis_c(option = "magma")


us_census %>% 
  ggplot() + 
  coord_sf() +
  aes(fips = FIPS) +
  stamp_sf_county_usa() +
  geom_sf_county_state(state = "new york") +
  aes(fill = mean_work_travel) + 
  scale_fill_viridis_c(option = "magma")
```

# Unify with Illinois

let's just make the original functions more flexible, and test with illinois

```{r}


compute_county <- function(data, scales, state = NULL, county = NULL){
 

  if(!is.null(state)){
    
  state %>% tolower() -> state
    
  reference_fips_full %>% 
                   mutate(name_2 = tolower(name_2)) %>% 
                   filter(name_2 %in% state) ->
  reference_fips_full
  }
    
  if(!is.null(county)){
    
    county %>% tolower() -> county
      
    reference_fips_full %>% 
                   mutate(name = tolower(name)) %>% 
                   filter(name %in% county) ->
    reference_fips_full
      
    }
    
    reference_fips_full %>% 
    select(fips, geometry, xmin, xmax, ymin, ymax) ->
reference_fips_state
  
  data %>% 
    inner_join(reference_fips_state) %>% 
    mutate(group = -1)

}


StatCounty <- ggplot2::ggproto(`_class` = "StatCounty",
                                  `_inherit` = ggplot2::Stat,
                                  # required_aes = c("fips"), #breaks when required!?
                                  # setup_data = my_setup_data,
                                  compute_panel = compute_county,
                                  default_aes = aes(geometry = after_stat(geometry))
                                  )



geom_sf_county <- function(
  mapping = NULL,
  data = NULL,
  position = "identity",
  na.rm = FALSE,
  show.legend = NA,
  inherit.aes = TRUE, ...) {
  ggplot2::layer(
    stat = StatCounty,  # proto object from step 2
    geom = ggplot2::GeomSf,  # inherit other behavior
    data = data,
    mapping = mapping,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(na.rm = na.rm, ...)
  )
}


us_census %>% 
  ggplot() + 
  coord_sf() +
  aes(fips = FIPS) +
  geom_sf_county(state = "Illinois") +
  geom_sf_county(state = "Illinois", 
                 county = c("Cook", "Champaign"), 
                 color = "red", linewidth = 1) +
  aes(fill = mean_work_travel) + 
  scale_fill_viridis_c(option = "magma")

last_plot() + 
  aes(fill = per_capita_income) + 
  scale_fill_viridis_c()

```

# Thank you Illinois... more 'tests'

```{r}

us_census %>% 
  ggplot() + 
  coord_sf() +
  aes(fips = FIPS) +
  geom_sf_county(state = "new york") +
  geom_sf_county(state = "new york", 
                 county = c("Putnam", "Dutchess", "Westchester", 
                            "Orange", "New Paltz", "Ulster"), 
                 color = "red", linewidth = 1) +
  aes(fill = mean_work_travel) + 
  scale_fill_viridis_c(option = "magma")


us_census %>% 
  ggplot() + 
  coord_sf() +
  aes(fips = FIPS) +
  geom_sf_county(state = c("California", "Oregon", 
                           "Washington", "Nevada", 
                           "Colorado", "Utah", "Arizona",
                           "New Mexico", "Texas", "New York", "Illinois")) +
  aes(fill = foreign_spoken_at_home) + 
  scale_fill_viridis_c()
  



```



---

# Notes

1. Inner join only seems to work in the compute panel move. 
- This may leave holes in your map.
- stamping down full map first may be a solution.
2. Finding GeomSf aesthetics behavior a bit surprising.  

- setting a required aesthetic seems to fail; like I'd require 'fips' as an aesthetic, 
- it seems to look for any declared aesthetic... I have to experience this problem again. I can't articulate it.  Now feels fuzzy.  Solution can be nulling it out.   Well, it was a surprise when ggplot2 was complaining about fips column not being found when using normal geom_sf() because it certainly never used the fips 'aesthetic' before.

3. I've found that using sf classically, produces xmin xmax ymin and ymax columns for the layer data.  They are required for my sf to render.  

4. Wishlist: Would be nice to have this plotly ready. I.e. convert using ggplotly and mouse over to see choropleth values and state and county names.  
